---
title: "LTER Data Sampler"
editor: visual
---

Sample Code from the [`lterdatasampler`](https://lter.github.io/lterdatasampler/index.html) package

# Cutthroat trout and salamander

length and weights in Mack Creek, Andrews Forest LTER

```{r}
#| label: fig-trout-salamander-size
#| fig-cap: Length and size of trout and salamander based on stream location.
#| warning: false
#| code-fold: true

library(tidyverse)
library(lterdatasampler)

trout_salamander <- and_vertebrates %>% 
  drop_na(species) %>% 
  filter(species != 'Cascade torrent salamander')

species_names <- c('Cutthroat trout' = 'cutthroat trout',
                   'Coastal giant salamander' = 'coastal giant salamander')

section_names <- c('CC' = 'clear cut forest',
                   'OG' = 'old growth coniferous forest')


ggplot(data = trout_salamander, aes(x = length_1_mm, y = weight_g)) +
  geom_point(aes(color = species), show.legend = FALSE) +
  labs(x = "Length (mm)",
       y = "Weight (g)",
       color = "Species") +
  facet_grid(section ~ species, labeller = labeller(section = section_names, species = species_names))
```

Now for some Python

```{python}
#| code-fold: true
import pandas as pd

trout_salamander_py = pd.read_csv('data/and_vertebrates.csv')
trout_salamander_py.head()
```

```{python}
trout_salamander_py = trout_salamander_py[trout_salamander_py['species'].isin(['Cutthroat trout', 'Coastal giant salamander'])]
```

```{python}
import matplotlib.pyplot as plt
trout_salamander_py.plot(x='length_1_mm', y='weight_g', kind='scatter')
plt.show()
```

# now to try tabset panels

::: callout-tip
## Tip

Use font awesome to add icons to the tab headers. To improve accessibility, use `a11y = "sem"` to get the icon title when you hover over it. Use `a11y = "deco"` for decorative icons and `a11y = "sem"` for semantic icons.

```markdown
::: panel-tabset
## 'r fontawesome::fa("r-project", fill = "#5A5A5A", a11y = "sem")` R

## 'r fontawesome::fa("python", fill = "#5A5A5A", a11y ="sem")` Python
:::
```

Does anyone know how to show the code above but with an actual backtick at the beginning of the lines?

```{markdown}
#| echo: fenced
::: panel-tabset
## `r fontawesome::fa("r-project", fill = "#5A5A5A", a11y = "sem")` R

## `r fontawesome::fa("python", fill = "#5A5A5A", a11y ="sem")` Python
:::
```
:::

::: panel-tabset
## `r fontawesome::fa("r-project", fill = "#5A5A5A", a11y = "sem")` R

```{r}
#| warning: false
ggplot(data = trout_salamander, aes(x = length_1_mm, y = weight_g)) +
  geom_point(aes(color = species), show.legend = FALSE) +
  labs(x = "Length (mm)",
       y = "Weight (g)",
       color = "Species") +
  facet_grid(section ~ species, labeller = labeller(section = section_names, species = species_names))
```

## `r fontawesome::fa("python", fill = "#5A5A5A", a11y ="sem")` Python

```{python}
trout_salamander_py = trout_salamander_py[trout_salamander_py['species'].isin(['Cutthroat trout', 'Coastal giant salamander'])]

import matplotlib.pyplot as plt
trout_salamander_py.plot(x='length_1_mm', y='weight_g', kind='scatter')
plt.show()
```
:::

# Citation

Horst A, Brun J (2022). lterdatasampler: Educational dataset examples from the Long Term Ecological Research program. R package version 0.1.0, <https://github.com/lter/lterdatasampler>.
